---
title: "ETO5513 Team Project"
author: "Gun Ho Jang & Rylie Douglas"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
# Load required libraries
library(tidyverse)
library(knitr)
library(kableExtra)
```

```{r load-data}
# Read the data
temp_data <- read.csv("~/Desktop/ETO5513A2/ETO5513Assessment2/Data/decadal-average-surface-temperature.csv")
```
# Introduction


## Section 1 - Macro Level - Gun Ho Jang


## Section 2 - Micro level - Rylie Douglas
### Research Questions
1. What are the patterns of temperature change across different climatic regions during this time period?

2. Which countries experienced the most unstable or fluctuating temperatures during this time period?

### Results and Analysis

#### Research Question 1: Temperature Comparisons Across Countries

```{r temperature-change, fig.width=12, fig.height=6}
# Calculate total temperature change for each country
temp_changes <- temp_data %>%
  group_by(Entity) %>%
  summarise(
    start_temp = first(Decadal.average.surface.temperature),
    end_temp = last(Decadal.average.surface.temperature),
    total_change = end_temp - start_temp
  ) %>%
  arrange(desc(abs(total_change)))

# Get top 10 countries with greatest change
top_10_countries <- head(temp_changes$Entity, 10)

# Create figure 1: Line graph for top 10 countries with enhanced width
temp_data %>%
  filter(Entity %in% top_10_countries) %>%
  ggplot(aes(x = Year, y = Decadal.average.surface.temperature, color = Entity)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  theme_minimal() +
  theme(
    legend.position = "right",
    legend.title = element_blank(),
    legend.text = element_text(size = 10),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold"),
    panel.grid.minor = element_blank(),
    plot.margin = margin(t = 20, r = 20, b = 20, l = 20)
  ) +
  labs(
    title = "Figure 1: Temperature Trends for Countries with Greatest Change (1940-2010)",
    x = "Year",
    y = "Average Temperature (°C)",
    caption = "Shows the 10 countries with largest absolute temperature change"
  ) +
  scale_x_continuous(breaks = unique(temp_data$Year)) +
  scale_y_continuous(breaks = seq(0, 30, by = 5)) +
  scale_color_brewer(palette = "Spectral")
```

```{r change-table}
# Table 1: Temperature changes for top 10 countries
temp_changes %>%
  head(10) %>%
  mutate(
    Rank = row_number(),
    total_change = round(total_change, 2),
    start_temp = round(start_temp, 2),
    end_temp = round(end_temp, 2)
  ) %>%
  select(Rank, Entity, start_temp, end_temp, total_change) %>%
  kable(
    caption = "Table 1: Countries with Greatest Temperature Change (1940-2010)",
    col.names = c("Rank", "Country", "1940 Mean Temperature (°C)", 
                  "2010 Mean Temperature (°C)", "Total Change (°C)")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```
#### Research Question 2: Temperature variability
```{r variability-analysis, fig.width=12, fig.height=6}
# Calculate temperature variability metrics
variability_analysis <- temp_data %>%
  group_by(Entity) %>%
  summarise(
    Temp_SD = sd(Decadal.average.surface.temperature),
    Temp_Range = max(Decadal.average.surface.temperature) - 
                 min(Decadal.average.surface.temperature),
    Temp_CV = (sd(Decadal.average.surface.temperature) / 
               mean(Decadal.average.surface.temperature)) * 100
  ) %>%
  arrange(desc(Temp_SD))

# Create figure 2: Top 15 most variable countries with enhanced width
ggplot(head(variability_analysis, 15), 
       aes(x = reorder(Entity, Temp_SD), y = Temp_SD)) +
  geom_bar(stat = "identity", fill = "#8B0000", alpha = 0.7) +
  coord_flip() +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold"),
    panel.grid.minor = element_blank(),
    plot.margin = margin(t = 20, r = 20, b = 20, l = 20)
  ) +
  labs(
    title = "Figure 2: 15 Countries with Highest Temperature Variability",
    x = "Country",
    y = "Temperature Standard Deviation (°C)",
    caption = "Higher values indicate greater temperature fluctuations"
  )
```
```{r variability-table}
# Table 2: Detailed variability metrics for top 10 countries
head(variability_analysis, 10) %>%
  mutate(
    Rank = row_number(),
    Temp_SD = round(Temp_SD, 2),
    Temp_Range = round(Temp_Range, 2),
    Temp_CV = round(Temp_CV, 2)
  ) %>%
  select(Rank, Entity, Temp_SD, Temp_Range, Temp_CV) %>%
  kable(
    caption = "Table 2: Top 10 Countries with Highest Temperature Variability",
    col.names = c("Rank", "Country", "Standard Deviation (°C)", 
                  "Temperature Range (°C)", "Coefficient of Variation (%)")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```


