---
title: "ETO5513 Team Project"
author: "Gun Ho Jang & Rylie Douglas"
date: "`r Sys.Date()`"
bibliography: references.bib    
output:
  html_document:
    toc: false
---

```{r setup, include=FALSE}
# Setting up
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.align = 'center')

# Load required libraries
library(tidyverse)
library(knitr)
library(kableExtra)
library(renv)
library(broom)

```

# Introduction


## Data Description
This analysis uses decadal surface temperature data (1940-2010) from Our World in Data enabling the analysis of long-term temperature patterns. The dataset spans from 1940 to 2010, measuring temperature changes across different countries globally through three main variables: country names, decadal time points, and mean surface temperatures in degrees Celsius (°C).

## Section 1 - Macro Level - Gun Ho Jang

The macro-level overview of the decadal average surface temperature globally will be addressed through the following questions:

1.  What is the global trend for average surface temperatures by decade (from 1940 to 2010)?
2.  Using regression analysis techniques, what is the expected temperature change in the coming decades?

### Research Question 1: What is the global trend for average surface temperatures by decade (from 1940 to 2010)?

The dataset is prepared for analysis by loading into a data frame.

```{r loading dataframe - Gun Ho part}
# Loading the 'decadal average surface temperatures' dataset into a data frame
decadal_avg_surface_temp <- read.csv("Data/decadal-average-surface-temperature.csv")
```

Next, the dataset is checked for any missing (N/A) values using the following code:

```{r missing values}
# Checking whether there are any missing values within the dataset
sum_na <- sum(is.na(decadal_avg_surface_temp))
print(paste("The dataset contains", sum_na, "N/A values."))
```

No missing values were found, so other data cleaning can be completed: the column headers are renamed, and the 'Code' column is removed as the Country column will be sufficient for country identification.

```{r data cleaning procedure}
# Renaming the columns and removing the 'Code' column
decadal_avg_surface_temp <- decadal_avg_surface_temp %>% 
  select(-Code) %>% 
  rename(
    `Country` = `Entity`,
    `Decadal Average Surface Temperature` = `Decadal.average.surface.temperature`
  )
```

The decadal trends between 1940 and 2010 for the global average surface temperature are summarised as follows:

```{r decadal trends- table}
# Grouping dataset by Year and calculating the average surface temperature 
decadal_trends <- decadal_avg_surface_temp %>% 
  group_by(Year) %>% 
  summarise(`Average Surface Temperature` = round(mean(`Decadal Average Surface Temperature`), 2))

# Representing decadal trends in a table
kable(
  decadal_trends,
  col.names = c("Decade", "Global Average Surface Temperature (°C)"),
  caption = "Table 1: Global Average Surface Temperature by Decade (1940-2010)"
)
```

A box-and-whisker plot visualisation was selected to represent the distribution of global average surface temperatures across the decades, and to highlight any outliers that appear in the dataset.

```{r decadal trends - graph, fig.width = 8, fig.height = 6}
# Plotting the global average surface temperatures over time in a box-and-whisker plot
ggplot(decadal_avg_surface_temp, aes(x = as.factor(`Year`), y = `Decadal Average Surface Temperature`)) +
  geom_boxplot() +
  stat_summary(
    fun = mean, 
    geom = "point",
    size = 2,
    color = "red"
    ) +
  stat_summary(
    fun = mean, 
    geom = "text",
    aes(label = round(after_stat(y), 2)),
    vjust = 2,
    color = "red"
    )  +
  theme_minimal() +
  labs(
    title = "Box-and-Whisker Plot of Average Surface Temperatures by Year",
    x = "Year",
    y = "Average Global Average Surface Temperature (°C)",
    caption = "The mean values have been included as red dot points."
  )
```

Both the table and the box-and-whisker plot demonstrate a steady increase in temperature from 1940 to 2010. The distribution shows an upward shift over the decades, with mean temperatures rising by 1.32°C during the observed period. Although this change appears small, the United Nations has warned that "every fraction of a degree of warming can result in discernible increases in the intensity and frequency of temperature and precipitation extremes" (@un_1_5c_matters).

The single low-temperature outlier in the plot corresponds with Greenland, reflecting the country's Arctic conditions. However, even it has also exhibited a concerning warming trend during the observed period.

### Research Question 2: Using regression analysis techniques, what is the expected temperature change in the coming decades?

A simple linear regression model was used to analyse the relationship between the temperature and time variables, enabling prediction of future trends based on the historical data.

```{r linear model - scatterplot, fig.width = 9, fig.height = 6}
# Constructing the linear model to observe the relationship between the two variables, Average Surface Temperature and Year
linear_model <- lm(`Decadal Average Surface Temperature` ~ Year, data = decadal_avg_surface_temp)

# Plotting a scatterplot with a linear regression line visualised
ggplot(decadal_avg_surface_temp, aes(x = `Year`, y = `Decadal Average Surface Temperature`)) +
  geom_line(
    aes(group = `Country`),
    color = "grey",
    alpha = 0.5
    ) +
  geom_smooth(
    method = "lm",  
    se = TRUE, 
    color = "red", 
    linetype = "dashed", 
    linewidth = 1.2
    ) +
  scale_x_continuous(
    breaks = seq(min(decadal_avg_surface_temp$Year), max(decadal_avg_surface_temp$Year), by = 10)
  ) +
  theme_minimal() +
  theme(
    plot.background = element_rect(fill = "white", color = NA)
  ) +
  labs(
    title = "Linear Regression on Decadal Average Surface Temperature Data",
    x = "Year",
    y = "Average Surface Temperature (°C)",
    caption = "Grey lines represent individual country trends while red line represents regression."
  )
```

The reliability of the model is assessed in the summary table below.

```{r linear model - summary table, echo = FALSE}
# Extracting coefficients and creating a summary table using tidy function
linear_model_summary <- tidy(linear_model) %>% 
  mutate(
    estimate = round(estimate, 4),
    std.error = round(std.error, 4),
    statistic = round(statistic, 4),
    p.value = round(p.value, 4)
  )

kable(linear_model_summary,
      col.names = c("Term", "Estimate", "Std. Error", "t-value", "p-value"),
      caption = "Linear Regression Model Summary"
      )
```

The summary table provides the following insights:

-   The slope (Year) indicates a temperature increase of 0.0187°C per year.

-   The Intercept has a p-value of 0.321, showing no statistical significance under null hypothesis testing. This is to be expected as it is not the focus of this analysis.

-   The Year variable's p-value of 0.0525 suggests a marginally significant relationship with the temperature. This relationship is only marginal as the p-value slightly exceeds the threshold of 0.05, weakly rejecting the null hypothesis at the 95% confidence level.

```{r predictions for future decades, echo = FALSE}
# Creating data frame to specify the future years for prediction
future_years <- data.frame(Year = c(2020, 2030, 2040, 2050))

# Predicting the temperatures for the next several decades using the linear model
future_years$Predicted_Temperature <- predict(linear_model, newdata = future_years)

# Rounding all predicted temperatures to 2 decimal places
future_years <- future_years %>% 
  mutate(Predicted_Temperature = round(Predicted_Temperature, 2))

# Tabulating the predicted outputs
kable(future_years,
      col.names = c("Year", "Predicted Temperature (°C)"),
      caption = "Predicted Temperatures for Next Several Decades"
      )
```

By 2050, the global average surface temperature is projected to reach approximately 19.40°C. However, the marginal relationship between year and temperature highlights the need for further analysis, including:

1.  Incorporating additional factors such as greenhouse gas emissions or the effects of ocean currents.

2.  Exploring non-linear modelling to capture complex temperature dynamics.

3.  Validating predictions with future observed data to improve the model's reliability.

## Section 2 - Micro level - Rylie Douglas

### Research Questions
This section investigates temperature variations at the country level, focusing on both long-term changes and temperature variability. The analysis addresses two primary research questions:

1. What are the patterns of temperature change across different climatic regions during this time period?

2. Which countries experienced the most unstable or variable temperatures during this time period?

### Results and Analysis

#### Research Question 1: Temperature Comparisons Across Countries

```{r load-data - Rylie part}
# Read the data
temp_data <- read.csv("Data/decadal-average-surface-temperature.csv")
```

```{r temperature-change, fig.width=12, fig.height=6}
# Calculate total temperature change for each country
temp_changes <- temp_data %>%
  group_by(Entity) %>%
  summarise(
    start_temp = first(Decadal.average.surface.temperature),
    end_temp = last(Decadal.average.surface.temperature),
    total_change = end_temp - start_temp
  ) %>%
  arrange(desc(abs(total_change)))

# Get top 10 countries with greatest change
top_10_countries <- head(temp_changes$Entity, 10)

# Create figure 1: Line graph for top 10 countries with enhanced width
temp_data %>%
  filter(Entity %in% top_10_countries) %>%
  ggplot(aes(x = Year, y = Decadal.average.surface.temperature, color = Entity)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  theme_minimal() +
  theme(
    legend.position = "right",
    legend.title = element_blank(),
    legend.text = element_text(size = 10),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold"),
    panel.grid.minor = element_blank(),
    plot.margin = margin(t = 20, r = 20, b = 20, l = 20),
    panel.background = element_rect(fill = "white", color = "black"),
    plot.background = element_rect(fill = "white", color = "black")
  ) +
  labs(
    title = "Figure 1: Temperature Trends for Countries with Greatest Change (1940-2010)",
    x = "Year",
    y = "Average Temperature (°C)",
    caption = "Shows the 10 countries with largest absolute temperature change"
  ) +
  scale_x_continuous(breaks = unique(temp_data$Year)) +
  scale_y_continuous(breaks = seq(0, 30, by = 5)) +
  scale_color_brewer(palette = "Spectral")
```
Figure 1 shows the trajectories of the ten countries with the most substantial temperature changes from 1940 to 2010. It can be seen that there is a slight upward trend of the yearly average  temperatures for these countries over time, revealing particularly notable increases in Nordic regions.

```{r change-table}
# Table 1: Temperature changes for top 10 countries
temp_changes %>%
  head(10) %>%
  mutate(
    Rank = row_number(),
    total_change = round(total_change, 2),
    start_temp = round(start_temp, 2),
    end_temp = round(end_temp, 2)
  ) %>%
  select(Rank, Entity, start_temp, end_temp, total_change) %>%
  kable(
    caption = "Table 1: Countries with Greatest Temperature Change (1940-2010)",
    col.names = c("Rank", "Country", "1940 Mean Temperature (°C)",
                  "2010 Mean Temperature (°C)", "Total Change (°C)")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "bordered"))
```
Table 1 reveals diverse patterns of temperature change across different climatic regions. Norway experienced the most significant increase of 2.91°C (from -1.70°C to 1.21°C), followed by Iceland with a 2.66°C rise (from -0.45°C to 2.21°C). Niger also shows a significant 2.65°C increase from a much higher average temperature (25.31°C to 27.96°C). It can be seen that cold climate countries (Norway, Iceland and Belarus) and hot climate countries (Niger, Saudi Arabia, Sudan, Eritrea) appear in the top rankings. Total mean temperature changes affected various climatic regions, however cold regions displayed slightly larger absolute changes over the 70-year period.

#### Research Question 2: Temperature variability
```{r variability-analysis, fig.width=12, fig.height=6}
# Calculate temperature variability metrics
variability_analysis <- temp_data %>%
  group_by(Entity) %>%
  summarise(
    Temp_SD = sd(Decadal.average.surface.temperature),
    Temp_Range = max(Decadal.average.surface.temperature) - 
                 min(Decadal.average.surface.temperature),
    Temp_CV = (sd(Decadal.average.surface.temperature) / 
               mean(Decadal.average.surface.temperature)) * 100
  ) %>%
  arrange(desc(Temp_SD))

# Create figure 2: Top 15 most variable countries with enhanced width
ggplot(head(variability_analysis, 15),
       aes(x = reorder(Entity, Temp_SD), y = Temp_SD)) +
  geom_bar(stat = "identity", fill = "#8B0000", alpha = 0.7) +
  coord_flip() +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold"),
    panel.grid.minor = element_blank(),
    plot.margin = margin(t = 20, r = 20, b = 20, l = 20),
    panel.background = element_rect(fill = "white", color = "black"),
    plot.background = element_rect(fill = "white", color = "black")
  ) +
  labs(
    title = "Figure 2: 15 Countries with Highest Temperature Variability",
    x = "Country",
    y = "Temperature Standard Deviation (°C)",
    caption = "Higher values indicate greater temperature fluctuations"
  )
```
Figure 2 displays temperature variability across 15 countries from 1940 to 2010, measured by standard deviation (SD). Nordic countries show the most significant fluctuations, with Norway recording the highest variability (SD = 1.17°C), followed closely by Iceland (SD = 1.10°C). This pattern aligns with previous studies of Scandinavian climate variability, which have shown that Nordic regions experience greater temperature fluctuations due to their atmospheric circulation patterns and geographical position [@hanssen2015climate]. These findings suggest that northern regions, particularly Scandinavia, experienced more unstable temperature patterns compared to other global regions during this period, which is consistent with research showing that high-latitude areas are more sensitive to climate variations.

```{r variability-table}
# Table 2: Detailed variability metrics for top 10 countries
head(variability_analysis, 10) %>%
  mutate(
    Rank = row_number(),
    Temp_SD = round(Temp_SD, 2),
    Temp_Range = round(Temp_Range, 2),
    Temp_CV = round(Temp_CV, 2)
  ) %>%
  select(Rank, Entity, Temp_SD, Temp_Range, Temp_CV) %>%
  kable(
    caption = "Table 2: Top 10 Countries with Highest Temperature Variability",
    col.names = c("Rank", "Country", "Standard Deviation (°C)",
                  "Temperature Range (°C)", "Coefficient of Variation (%)")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "bordered"))
```
Table 2 presents temperature variability through three distinct metrics for the top 10 countries: standard deviation measuring absolute variability, temperature range indicating the interval between the extremes and coefficient of variation (CV) indicating relative variability. Norway has the highest standard deviation (1.17°C), temperature range (3.07°C) and has a negative coefficient of variation (-290.82%), which suggests extreme temperature fluctuations around its mean. Iceland shows the next highest CV  of 168.34%. On the other hand, other high ranking countries like Saudi Arabia, Niger, and Yemen exhibit more moderate coefficients between 2-4%, this indicates that their temperature variations were less extreme relative to their higher mean temperatures.

# Conclusion

# References

